# Module 5

<!-- Things to Tell AI:
1. You want answers stored locally when the person downloads the book, 
2. When reset you want to do a hard reset (so wipes it even if they move pages
3. Do each question and part separately
4. Hint, check and reset buttons like Ive done in Mod 1
-->

<!-- ====== COLOURED BANNER ====== -->
```{=html}
<div style="
  background: linear-gradient(90deg, #e6f0ff, #f7fbff);
  border-left: 8px solid #2a7ae2;
  padding: 22px;
  margin-bottom: 30px;
  border-radius: 6px;
">
  <h2 style="margin-top:0; color:#1a4f8b; font-weight:700;">
    Module Focus:
  </h2>
  <p style="font-size:1.05em; margin-bottom:0;">
    One line summary

In this module, we ???.
  </p>
</div>
```
<!-- ====== END BANNER ====== -->

## Summary

## Supplementary Notes

De Veaux, Velleman & Bock’s approach to probability is at times more mathematical than necessary for this course, so we have prepared these notes to supplement explanation of the important concepts and to highlight where some content in the relevant chapters is not required.

### Introduction
The concept of probability is familiar from everyday events and experiences. We use the language of probability in all sorts of contexts ranging from describing the vagaries of the weather to the anticipated results of sporting contests and expectations about our own future. Words such as likely, probable, unexpected, impossible and so on are part of our language.

Although we don’t usually associate a specific number with such descriptions we are aware that outcomes which are unlikely have a probability near 0% and outcomes which are nearly certain have a probability near 100%. The simple idea of using numbers in the range 0 to 1 or 0% to 100% to quantify probabilities leads to many useful results. For example, if we are 100% certain of dying then we must be 0% certain of living forever. If a horse is 10% certain of winning a race it must be 90% certain of losing. In general terms the probability of an event occurring is one minus the probability of it not occurring. This is a rule of probability.

This module formalises these familiar probability properties and rules. The second half of the module introduces the underlying principles of statistical inference – hypothesis testing and confidence intervals. The remaining modules in this course include a variety of hypothesis testing contexts and related statistical concepts all of which have their foundations in probability theory.


## Randomness and Probability
Answers to probability questions can be expressed as a decimal fraction (such as 0.28) or a percentage (such as 28%). As usual, rounding off the final answer sensibly is recommended.
Knowing the properties and rules about probabilities is one thing; being able to solve problems using them is another. The only way to get good at it is to practise enough problems to recognise what’s required. There is a minimum list of problems to at tempt in the corresponding tutorial in the Workbook. If you find them straightfor ward, great. If not, do some more now or later.
If you enjoy working with probabilities (some of us do!), read De Veaux, Velleman & Bock, (4th edition), Chapter 14 (De Veaux, Velleman & Bock, (5th edition), Chapter 13). None of that material is assessable however!

```{r, fig.width=5, fig.height=5, out.width="60%", fig.cap="Probability distribution for number of heads in 3 tosses", echo=FALSE}
x <- 0:3
prob <- dbinom(x, size = 3, prob = 0.5)

plot(x, prob, type="h", lwd=5, col="blue",
     xlab="x, number of heads",
     ylab="probability",
     ylim=c(0,0.4))

points(x, prob, pch=16, col="blue")
```

The probability distribution for the number of heads when tossing a fair coin three times is:

\[
P(X = x) = \binom{3}{x} (0.5)^3,\quad x = 0,1,2,3.
\]

The standard deviation is:

\[
\sigma = \sqrt{
0.125(0 - 1.50)^2 +
0.375(1 - 1.50)^2 +
0.375(2 - 1.50)^2 +
0.125(3 - 1.50)^2
}
\]

\[
\sigma = \sqrt{0.75} = 0.866.
\]

# Review Questions: Module 5 {-}
<!-- Copy from Module 1. Then copy from the Workbook and ask AI to change  -->

<!--
NOTE FOR FUTURE ME:
This module uses child Rmd files under section headings (##) so that:
- Module 1, Tutorial are in different files but render together
- Summary, Review, and Tutorial are numbered 1.1, 1.2, 1.3
- bs4_book splits these sections into separate HTML pages (via split_by: section)
- PDF output still renders cleanly with correct numbering

DO NOT add this file's children to _bookdown.yml
DO NOT use # headings in the child files
-->

```{r, child="05-Tut5.Rmd"}
```

